# Web-scraping

## Introduction

Getting content off websites can be a nightmare. The worst case resort is manually typing data from a web-page into spreadsheets... but there are ***many*** steps we can do before resorting to that.

This chapter will outline the process for pulling data off the web, and particularly for understanding the exact web-page element we want to extract. The notes and code loosely follow the fabulous tutorial by Grant R. McDermott in his [*Data Science for Economists*](https://raw.githack.com/uo-ec607/lectures/master/06-web-css/06-web-css.html)series.

First up, let's load some packages.

```{r,results='hide', warning=FALSE,message=FALSE}

# Load and install the packages
pacman::p_load(tidyverse, dplyr, rvest, lubridate, janitor, data.table, hrbrthemes)
```

## Anatomy of a webpage

Let's introduce some terminology: ***server side***

This describes information being embeded directly in the webpage's HTML. When dealing with server side webpages, finding the correct CSS (or Xpath) "selectors" becomes the hardest part of the task.

Iterating through dynamic webpages (e.g. "Next page" and "Show More" tabs) is also tricky - but we'll get there.

Trawling through CSS code on a webpage is a bit of a nightmare - so we'll use a chrome extension called [SelectGadget](https://chrome.google.com/webstore/detail/selectorgadget/mhjhnkcfbdhnjickkkdbjoemdmbfginb) to help.

The R package that's going to do the heavy lifting is called [`rvest`](https://github.com/tidyverse/rvest) and is based on the python package called Beauty Soup.

## Scraping a table

Let's use this [wikipedia page](https://en.wikipedia.org/wiki/Men%27s_100_metres_world_record_progression#Unofficial_progression_before_the_IAAF) as a starting example. It contains various entries for the men's 100m running record.

We can start by pulling all the data from the webpage.

```{r, warning=FALSE,message=FALSE}

m100 <- rvest:: read_html("https://en.wikipedia.org/wiki/Men%27s_100_metres_world_record_progression") 
m100

```

...and we get a whole heap of mumbo jumbo.

To get the table of 'Unofficial progression before the IAAF' we're going to have to be more specific.

Using the SelectGadget tool we can click around and identify that that specific table is named in the HTML code: **.wikitable :nth-child(1)**

Side note: The names of tables in HTML webpages changes all the time. Therefore this code is considered unstable in terms of setting and forgetting.

```{r, warning=FALSE,message=FALSE}

pre_iaaf <- 
  m100 %>%
  #Select table
  rvest::html_element(".wikitable :nth-child(1)") %>%
  #Convert to data frame
  rvest::html_table()                                     

pre_iaaf

```

Niiiiice - now that's better. Let's do some quick data cleaning.

```{r,warning=FALSE,message=FALSE}

pre_iaaf <- pre_iaaf %>%
            clean_names() %>%
            dplyr::mutate(date = mdy(date))
  
pre_iaaf
```

Let's also scrape the data for the more recent running records. That's the tables called 'Pre-automatic timing (1912--1976)', and 'Modern Era (1977 onwards)'.

For the second table:

```{r,warning=FALSE,message=FALSE}

iaaf_76 <- m100 %>%
           html_element("h3+ .wikitable") %>%
           rvest::html_table()

iaaf_76 <- iaaf_76 %>%
           clean_names() %>%
           mutate(date = mdy(date)) 

iaaf_76

```

And now for the third table:

```{r,warning=FALSE,message=FALSE}

iaaf <- m100 %>%
        html_element("p+ .wikitable") %>%
        html_table() %>%
        clean_names() %>%
        mutate(date = mdy(date))
iaaf

```

How good. Now let's bind the rows together to make a master data set.

```{r,warning=FALSE,message=FALSE}

wr100 <- rbind(
    pre_iaaf %>% dplyr::select(time, athlete, nationality, date) %>% 
    mutate(era = "Pre-IAAF"),
    iaaf_76 %>% dplyr::select(time, athlete, nationality, date) %>% 
    mutate(era = "Pre-automatic"),
    iaaf %>% dplyr::select(time, athlete, nationality, date) %>% 
    mutate(era = "Modern")
    )

wr100

```

Excellent. Let's plot the results.

```{r,warning=FALSE,message=FALSE}

ggplot(wr100)+
  geom_point(aes(x = date, y = time, col = era),alpha=0.7)+

  labs(title="Men's 100m world record progression",
  subtitle = "Analysing how times have improved over the past 130 years",
  caption = "Data: Wikipedia 2021",
  x="",
  y="") + 
  
  theme_minimal()+
  
  scale_y_continuous(limits=c(9.5,11), breaks=c(9.5,10,10.5,11))+
  
  theme(axis.text.y = element_text(vjust = -0.5, margin = ggplot2::margin(l = 20, r = -20)))+
  
  theme(plot.subtitle = element_text(margin=ggplot2::margin(0,0,25,0))) +
  
  theme(legend.title = element_blank())+
  
  theme(plot.title=element_text(face="bold",size=12))+
  theme(plot.subtitle=element_text(size=11))+
  theme(plot.caption=element_text(size=8))+
  
  theme(axis.text=element_text(size=8))+
  theme(panel.grid.minor = element_blank())+
  theme(panel.grid.major.x = element_blank()) +
  
  theme(axis.line.x = element_line(colour ="black",size=0.4))+
  theme(axis.ticks.x = element_line(colour ="black",size=0.4))
```
