---

---

# Economic indicators

## Overview

Australia has exceptional financial and economic institutions. Three of these institutions release periodic data useful for economic analysis:

-   [Australian Bureau of Statistics](https://www.abs.gov.au/)

-   [Reserve Bank of Australia](https://www.rba.gov.au/)

-   [Treasury](https://treasury.gov.au/)

As usual, there are catches. Most of this data is in inconsistent formats (the reasons for which continue to baffle me). What's more, it's currently not possible to ping databases or API's for access to this data... it is mainly accessed through spreadsheets.

The scripts below run through some of the main ways to import, clean, and analyse Australian macroeconomic data in R.

Some of the key packages we'll use are [readabs](https://github.com/MattCowgill/readabs) and [readrba](https://github.com/MattCowgill/readrba).

To get started, let's install and load packages.

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

#Loads the required required packages
pacman::p_load(ggmap, ggplot2, dplyr, tmaptools, RCurl, jsonlite, tidyverse, leaflet, writexl, readr, readxl, readabs, readrba,lubridate, zoo, scales)
```

## Gross Domestic Product

To get GDP data from the ABS, we'll use the read_abs function from the `readrba` package.

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

#For simplicity, we keep the download function seperate to the analysis
all_gdp <- read_abs("5206.0")
```

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

#Select the seasonally adjusted data and filter for data and value columns
gdp_level <- all_gdp %>% 
              filter(series == "Gross domestic product: Chain volume measures ;", 
                     !is.na(value)) %>%
              filter(series_type =="Seasonally Adjusted") %>%
              dplyr::select(date,value) %>%
              dplyr::rename(quarterly_output=value)

gdp_level <- gdp_level %>%
             mutate(quarterly_growth_rate = 
                    ((quarterly_output / lag(quarterly_output,1)-1))*100) %>%
             mutate(annual_gdp =
                    rollapply(quarterly_output, 
                              4, 
                              sum, 
                              na.rm=TRUE, 
                              fill = NA, 
                              align = "right")) %>%
             mutate(annual_gdp_trillions=annual_gdp/1000000)%>%
             mutate(annual_growth_rate = 
                   ((annual_gdp / lag(annual_gdp, 4) - 1))*100)%>%
             mutate(Quarter_of_year =
                    lubridate::quarter(date,
                                       with_year = FALSE, 
                                       fiscal_start = 1))

#Set a baseline value
gdp_level$baseline_value <- gdp_level$quarterly_output[
                            which(gdp_level$date =="2022-03-01")]


gdp_level <- gdp_level %>%
             mutate(baseline_comparison =
                   (quarterly_output/baseline_value)*100)

tail(gdp_level)
```

Now we can plot the GDP data for Australia.

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

plot_gdp <- ggplot(data=gdp_level)+
            geom_line((aes(x=date, y=annual_gdp_trillions)), col="blue") +
  
            labs(title = "Australian GDP ($AUD)",
                 subtitle = "Annualised figures",
                 caption = "Data: Australian Bureau of Statistics",
                 y = "", 
                 x = " ")+
  
            scale_y_continuous(breaks = c(0,0.5,1.0,1.5,2.0,2.5),
                               labels = label_number(suffix = " trillion"))+
  
            scale_x_date(date_breaks = "10 years", date_labels="%Y")+
  
            theme_minimal() +
            theme(legend.position="bottom")+
  
            theme(plot.title=element_text(face="bold", size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))+

            theme(axis.text=element_text(size=8))+
            theme(panel.grid.minor = element_blank())+
            theme(panel.grid.major.x = element_blank()) +

            theme(axis.title.y = 
            element_text(margin = margin(t = 0, r = 0, b = 0, l = 0)))+
                         
            theme(axis.text.y = element_text(vjust = -0.5, 
                                             margin = margin(l = 20, r = -45)))+
  
            theme(axis.line.x = element_line(colour ="black", size=0.4))+
            theme(axis.ticks.x = element_line(colour ="black", size=0.4))


plot_gdp
```

## Unemployment rate

Download the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

#Download the time sreies
all_unemployment <- read_abs("6202.0")

```

Clean and analyse the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

unemployment_rate <- all_unemployment %>% 
              filter(series == "Unemployment rate ;  Persons ;",!is.na(value)) %>%
              filter(table_title=="Table 1. Labour force status by Sex, Australia - Trend, Seasonally adjusted and Original") %>%
              filter(series_type =="Seasonally Adjusted") %>%
              mutate(mean_unemployment_rate=mean(value)) %>%
              mutate(percentile_25=quantile(value,0.25))%>%
              mutate(percentile_75=quantile(value,0.75)) %>%
              dplyr::select(date,value,mean_unemployment_rate,percentile_25,percentile_75)
              

tail(unemployment_rate)
```

Plot the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

plot_unemployment_rate <- ggplot(data=unemployment_rate)+
                          geom_line(aes(x = date, y = value), col = "blue") +

            labs(title = "Unemployment rate",
                 subtitle = "Subtitle goes here",
                 caption = "Data: Australian Bureau of Statistics",
                 y = "Unemployment rate (%)", 
                 x = " ")+
  
            scale_y_continuous(labels = scales::comma)+
            scale_x_date(date_breaks = "10 years", date_labels="%Y")+
  
            theme_minimal() +
            theme(legend.position="bottom")+
  
            theme(plot.title=element_text(face="bold", size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))+

            theme(axis.text=element_text(size=8))+
            theme(panel.grid.minor = element_blank())+
            theme(panel.grid.major.x = element_blank()) +

            theme(axis.title.y = 
            element_text(margin = margin(t = 0, r = 0, b = 0, l = 0)))+
                         
            theme(axis.text.y = element_text(vjust = -0.5, 
                                             margin = margin(l = 20, r = -15)))+
  
            theme(axis.line.x = element_line(colour ="black", size=0.4))+
            theme(axis.ticks.x = element_line(colour ="black", size=0.4))

plot_unemployment_rate

```

## Inflation (CPI)

Download the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

all_CPI <- read_abs("6401.0")

```

Clean and analyse the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

Australia_CPI <- all_CPI %>% 
                 filter(series == "Percentage Change from Corresponding Quarter of Previous Year ;  All groups CPI ;  Australia ;",!is.na(value)) %>%
                 mutate(mean_CPI=mean(value)) %>%
                 mutate(percentile_25=quantile(value,0.25))%>%
                 mutate(percentile_75=quantile(value,0.75)) %>%
                 dplyr::select(date, value,mean_CPI,percentile_25,percentile_75)
              
tail(Australia_CPI)

#Can add in the below line to filter
#filter(date>"2010-01-01") %>%
```

Plot the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

plot_CPI <- ggplot(data=Australia_CPI %>% 
                   filter(date>(as.Date("2000-01-01"))))+
                          
                   geom_rect(aes(xmin=as.Date("2000-01-01"), 
                                 xmax=as.Date("2023-03-01"), 
                                 ymin=2, 
                                 ymax=3),
                                 alpha=0.01,
                                 fill="grey")+
                   geom_line(aes(x = date, y = value), col = "blue") +
  
                   scale_x_continuous(expand=c(0,0))+
  
            labs(title = "Inflation (as measured by the CPI)",
                           subtitle = "Subtitle goes here",
                           caption = "Data: Australian Bureau of Statistics",
                           y = "(%)", 
                           x = " ")+
  
            scale_y_continuous(labels = scales::comma)+
            scale_x_date(date_breaks = "5 years", date_labels="%Y")+
  
            theme_minimal() +
            theme(legend.position="bottom")+
  
            theme(plot.title=element_text(face="bold", size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))+

            theme(axis.text=element_text(size=8))+
            theme(panel.grid.minor = element_blank())+
            theme(panel.grid.major.x = element_blank()) +

            theme(axis.title.y = 
            element_text(margin = margin(t = 0, r = 0, b = 0, l = 0)))+
                         
            theme(axis.text.y = element_text(vjust = -0.5, 
                                             margin = margin(l = 20, r = -15)))+
  
            theme(axis.line.x = element_line(colour ="black", size=0.4))+
            theme(axis.ticks.x = element_line(colour ="black", size=0.4))

plot_CPI

```

Plot a histogram of the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

plot_CPI_hist <- ggplot(Australia_CPI, aes(x=value)) + 
                        geom_histogram(aes(y=..density..), 
                                       colour="black", fill="lightblue")+
  
                        geom_density(alpha=.5, fill="grey",colour="darkblue")+
  
                        scale_x_continuous(expand=c(0,0))+
  
            labs(title = "Consumer Price Index: Histogram",
                           subtitle = "Subtitle goes here",
                           caption = "Data: Australian Bureau of Statistics",
                           y = "(%)", 
                           x = " ")+
  
            theme_minimal() +
            theme(legend.position="bottom")+
  
            theme(plot.title=element_text(face="bold", size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))+

            theme(axis.text=element_text(size=8))+
            theme(panel.grid.minor = element_blank())+
            theme(panel.grid.major.x = element_blank()) +

            theme(axis.title.y = 
            element_text(margin = margin(t = 0, r = 0, b = 20, l = 0)))+
                         
            theme(axis.text.y = element_text(vjust = -0.5, 
                                             margin = margin(l = 20, r = -2)))+
  
            theme(axis.line.x = element_line(colour ="black", size=0.4))+
            theme(axis.ticks.x = element_line(colour ="black", size=0.4))


plot_CPI_hist
```

## Wage Price Index

Download the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

all_wpi <- read_abs("6345.0")

```

Clean and analyse the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

Australia_WPI <- all_wpi %>% 
                 filter(series == "Percentage Change From Corresponding Quarter of Previous Year ;  Australia ;  Total hourly rates of pay excluding bonuses ;  Private and Public ;  All industries ;",
                                  !is.na(value)) %>%
                 filter(series_type=="Seasonally Adjusted") %>%
                 mutate(mean_WPI=mean(value)) %>%
                 dplyr::select(date, value,mean_WPI)

tail(Australia_WPI)
```

Plot the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

plot_WPI <- ggplot(data=Australia_WPI)+
                   geom_line(aes(x = date, y = value), col = "blue") +

            labs(title = "Wage Price Index",
                           subtitle = "Subtitle goes here",
                           caption = "Data: Australian Bureau of Statistics",
                           y = "(%)", 
                           x = " ")+
  
            scale_y_continuous(labels = scales::comma)+
            scale_x_date(date_breaks = "5 years", date_labels="%Y")+
  
            theme_minimal() +
            theme(legend.position="bottom")+
  
            theme(plot.title=element_text(face="bold", size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))+

            theme(axis.text=element_text(size=8))+
            theme(panel.grid.minor = element_blank())+
            theme(panel.grid.major.x = element_blank()) +

            theme(axis.title.y = 
            element_text(margin = margin(t = 0, r = 0, b = 0, l = 0)))+
                         
            theme(axis.text.y = element_text(vjust = -0.5, 
                                             margin = margin(l = 20, r = -15)))+
  
            theme(axis.line.x = element_line(colour ="black", size=0.4))+
            theme(axis.ticks.x = element_line(colour ="black", size=0.4))
                          

plot_WPI
```

## RBA cash rate

Download the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

cash_rate_all<- readrba::read_rba(table_no = "F13")
head(cash_rate_all)
```

Clean and analyse the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

cash_rate_Australia <- cash_rate_all %>%
                       filter(series=="Australia Target Cash Rate") %>%
                       dplyr::select(date, value)

tail(cash_rate_Australia)
```

Plot the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

plot_cash_rate <- ggplot(data=cash_rate_Australia)+
                          geom_line(aes(x = date, y = value), col = "blue") +
  
            labs(title = "RBA cash rate",
                           subtitle = "Subtitle goes here",
                           caption = "Data: Read RBA",
                           y = " ", 
                           x = " ")+
  
            scale_y_continuous(labels = scales::comma)+
            scale_x_date(date_breaks = "3 years", date_labels="%Y")+
  
            theme_minimal() +
            theme(legend.position="bottom")+
  
            theme(plot.title=element_text(face="bold", size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))+

            theme(axis.text=element_text(size=8))+
            theme(panel.grid.minor = element_blank())+
            theme(panel.grid.major.x = element_blank()) +

            theme(axis.title.y = 
            element_text(margin = margin(t = 0, r = 0, b = 0, l = 0)))+
                         
            theme(axis.text.y = element_text(vjust = -0.5, 
                                             margin = margin(l = 20, r = -15)))+
  
            theme(axis.line.x = element_line(colour ="black", size=0.4))+
            theme(axis.ticks.x = element_line(colour ="black", size=0.4))
                          
plot_cash_rate

```

## AUD exchange rate

*God knows why - but but there are super quirky names for the official exchange rate tables*

Download the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

exchange_rate_all<- readrba::read_rba(table_no = (c("ex_daily_8386",
                                                    "ex_daily_8790",
                                                    "ex_daily_9194",
                                                    "ex_daily_9598",
                                                    "ex_daily_9902",
                                                    "ex_daily_0306",
                                                    "ex_daily_0709",
                                                    "ex_daily_1013",
                                                    "ex_daily_1417",
                                                    "ex_daily_18cur")),
                                      cur_hist = "historical")

```

Clean and analyse the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

exchange_rate_AUD <-   exchange_rate_all %>%
                       filter(series=="A$1=USD") %>%
                       dplyr::select(date, value)


tail(exchange_rate_AUD)

```

Plot the data

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.keep = TRUE}

plot_exchange_rate_AUD <- ggplot(data=exchange_rate_AUD)+
                          geom_line(aes(x = date, y = value), col = "blue") +

            labs(title = "AUD exchange rate",
                           subtitle = "Subtitle goes here",
                           caption = "Data: Reserve Bank of Australia",
                           y = " ", 
                           x = " ")+
  
            scale_y_continuous(labels = scales::comma)+
            scale_x_date(date_breaks = "3 years", date_labels="%Y")+
  
            theme_minimal() +
            theme(legend.position="bottom")+
  
            theme(plot.title=element_text(face="bold", size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))+

            theme(axis.text=element_text(size=8))+
            theme(panel.grid.minor = element_blank())+
            theme(panel.grid.major.x = element_blank()) +

            theme(axis.title.y = 
            element_text(margin = margin(t = 0, r = 0, b = 0, l = 0)))+
                         
            theme(axis.text.y = element_text(vjust = -0.5, 
                                             margin = margin(l = 20, r = -15)))+
  
            theme(axis.line.x = element_line(colour ="black", size=0.4))+
            theme(axis.ticks.x = element_line(colour ="black", size=0.4))


plot_exchange_rate_AUD

```
